The yaq Project: a Protocol for Scientific Instrumentation

### 2022-05-13 paper.tex
lots of content

@@

\subsection{Configuration and State}

@@

\subsection{Properties and Traits}

Daemon roundup.

Discussion of traits and purpose of compositional approach.

\subsection{Traits}

Graphic of traits?

Examples of daemons for core traits.

Fakeable.

@@

Introductory paragraph.

The Landis Group at UW-Madison created WiQK, a quench kinetics reactor.
Reactions can be orchestrated with a simple script using Python's built-in time library and yaqc.
This has worked for a few years as developers continue to modify their reactor.

The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
From an automation control perspective, this reactor incorporates a collection of sensors and a single heating process value under computer control
This reactor was made as a self-contained unit with a Raspberry Pi [CITE] as the computer.
Users interact with a custom Qt-based GUI to use the reactor.
yaq daemons run locally along-side the lightweight Qt application making the entire system simple to run.

The Wright Group needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments.
We have built out a toolchain incorporating yaq into Bluesky to take advantage of the sophisticated orchestration that Bluesky supports.
We are excited about Bluesky moving forward.

>>>>

In yaq, each component of an instrument is supported by a tiny lightweight process that runs in the background of your computer: a daemon.
Since each component has its own process, each daemon can be developed separately.
Using yaq, a typical instrument might have several daemons each supporting a particular component.
Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
To do an experiment, a control program (a "client") must send commands to each of these daemons.
Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.

The separation between each daemon and client makes the yaq framework less fragile than monolithic applications.
The yaq framework is language agnostic.
For example, a daemon written in Python might be controlled by a client implemented using LabVIEW.
In yaq, each component of an instrument can be developed and distributed separately.
For example, two very different instruments might happen to use the same temperature sensor.
Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
As yaq grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.

There are currently 62 daemons in the yaq project supporting at least 40 kinds of hardware.
Because yaq is protocol based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the yaq website.

@@

@@

Properties and traits in general.

\section{Traits}

Each component of a scientific instrument may have many parameters.
Similar components typically have similar parameters, but there may be incidental differences between each implementation.
For example, consider two different monochromators.
The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.

yaq defines "traits": collections of behaviors which are logically grouped together.
Each trait specifies a set of configuration options, state entries, and methods that a daemon can implement.
Daemons, then, can be composed of multiple traits.
Clients can trust that daemons which implement a given trait will behave in similar ways.
For example, any daemon that implements the "is-position" trait must expose the methods "get\_position" and "set\_position".
It becomes easy to write generic clients that work with any daemon that has certain traits.
Of course, daemons are always allowed to implement additional configuration, state, and methods in addition to those which are implied by their traits.

@@

yaq's simplicity and extensible nature has allowed us to use it in a broad variety of instrumentation designs.
We highlight a few such designs here.

The Landis Group at UW-Madison is currently working on a new type of flow reactor: the Wisconsin Quench Kinetics Reactor (WiQK).
This reactor incorporates several computer-controlled valves and syringe pumps as well as various sensors.
The WiQK is rapidly evolving as researchers continue to test and refine their design.
Only a few researchers are actively using the reactor during this prototyping stage.
The Landis Group has written very basic Python scripts utilizing yaqc to orchestrate hardware for their reactor.
These lightweight scripts can be extensively refactored as the hardware and orchestration strategy changes dramatically during WiQK development.
This approach ensures that the Landis Group is not limited by their ability to orchestrate hardware as they refine their reactor design.

The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
yaq daemons are used to interface with each sensor and the heater controller.
We were able to create a lightweight Qt-based GUI which offers an easy user experience to researchers.
By separating hardware interface code from GUI code, each piece of the system became easier to produce and maintain.
Taken together, these components have been a solid software experience for many users over several years.

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using yaq, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.

yaq is also being used in several smaller ways throughout UW-Madison Chemistry.
Excitingly, all of these research groups are able easily benefit from each-other's daemon developments.
This level of collaboration is new to us in the orchestration software space.
